Revealing transient strain in geodetic data with Gaussian process regression

### 2017-11-10 ResponseToReviewers/ResponseToReviewers.tex

\textbf{1- Now, the correlation of the stochastic processes is not completely
clear. When the authors justify the model behind equation 3, wij is a
normally distributed, uncorrelated noise. In statistic, it is equal to
a WGN process. Then, the authors introduce a separate parameter (eta)
which models the temporally correlated noise. This parameter is also
following a Gaussian distribution, with a zero-mean and a covariance
matrix $C_{\eta}$ (i.e. line 10 p5). This approach is not common in
modelling GNSS time series. The standard is based on Williams 2003,
where the author established a stochastic model with a covariance
matrix as sum of identity matrix for the white noise and another
matrix which represents the coloured noise. The coloured noise
covariance matrix is defined differently (see the literature – Bock and
Melgar Reviews of Geophysics 2016, He et al., Journal of geodynamics
2017).}

I have changed my description of the data (eq. 3 in the revised draft)
so that it includes a single term, $\eta$, that encompasses correlated
and uncorrelated noise. I then break $\eta$ down into correlated and
uncorrelated components in Section 3.1 (eq. 25). The covariance matrix
for $\eta$ in eq. 26 is more consistent with the literature.

\textbf{2- First, the author should give a summary of stochastic noise
modelling in GNSS. I also think that a discussion is needed to relate
to previous models. Why using separated stochastic models? Also I
would comment on the possibility that the (low) spatial correlation
between the parameters in your deterministic model (i.e. secular
velocities) could be absorbed in the estimation of the covariance
matrix $C_{\eta}$.}

I believe the reviewer is refering to page 4-5 of the original draft,
where I first introduce $\eta$. I agree that a more thorough
discussion on the noise model is necessary, but I do not think this is
an appropriate place in the text for it. The paper is laid out so that
Section 2 contains a general description of the method for estimating
transient strain rates, and Section 3 contains details that are
specific to the particular application. There is no universally
appropriate noise model, so I decided to save a discussion on the
noise model for Section 3. This is clarified in Section 2 of the
revised draft when I say "The appropriate noise model may vary
depending on the application, and we hold off on specifying the
covariance matrix, $C_{\eta_i}$, until Section 3.1."

@@

\textbf{Now, I am confused When looking at p.8 Section 4.1, the
description of the noise models should be given as a subsection of
Section 2 p.4. I would also improve the literature review. Several
other models have also been discussed (i.e. Montillet et al., 2014
uses a fractional Brownian motion model; ARMA, ARFIMA, GGM or Band
pass noise – see Het et al., 2017).}

Again, we think that the description of the noise model should remain
in Section 3 because it is specific to the Pacific Northwest
application.

We have added a few more references in Section 3.1 of the revised
draft TODO to emphasize that GNSS noise modeling is a well-worked
area. We refer the reader to Bock and Melgar 2016 and He et al 2017
for a more complete review.

@@

\textbf{Outlier detection:}

\textbf{1- P7 : I would give a longer summary in the introduction about all
the efforts in outlier detection in GNSS time series. The first
paragraph p.7 is not enough.}

In Section 2.2 of the revised draft, I have expanded my discussion on
how previous studies have treated outliers in GNSS data.

@@

\textbf{Results:}

\textbf{The results are generally well explained. However, the use of
different priors and the comparison in the text is sometimes
confusing. Perhaps, a table could summary the different priors and the
main results. It would ease the reading of Section 4.2, 4.3 and then
the discussion in Section 5.}

In the previous draft, I was discussing two different priors: a prior
for the outlier detection algorithm, and a prior used in computing
transient strain rates. I acknowledge that this was convoluted. In the
revised draft, all the details of the outlier detection algorithm are
confined to the appendix. I hope this clarifies any confusion.

Also, I have rewritten most of the results section to be more
structured and hopefully have a clearer focus.

@@

\textbf{P2 line 30 : “fidelity” replace with “reliability”.}

Done

\textbf{P2 line 30-31 “ Developing and improving upon methods for deriving
secular...area of research.” Need references.}

The introduction has been revised and this sentence has been removed.

@@

\textbf{P.20 : “ Using a compact covariance function ..” Again I would just
say that the covariance matrix is sparse and that s why you have the
mathematical simplification.}

My personal preference is to leave the wording as is.

\textbf{p.20 line 33 “ computational burden” I think that refers to p.20 line
60 “prohibitive when using sveral years of daily GNSS”. It would
interesting to have some number. How do you quantify the computation
time to process longer and longer time series?}

I agree that numerical values would be useful; however, I am reluctant
to quantify the computational cost of evaluation transient strain
rates. There are many factors that influence the computational cost
(whether sparse matrices are used, the type of sparse matrices that
are used, the linear solver algorithm, etc.) and I am not convinced
that my implementation is the most efficient. So I do not give details
on the computation time, but I note in the discussion that this is an
area for further research.

### 2017-11-10 RevisedSubmission/hines_hetland_2017_gji.tex
did more

% Why we care about determining strain

Crustal strain rates are fundamentally important quantities for
assessing seismic hazard. Knowing where and how quickly strain is
accumulating gives insight into where we can expect stored elastic
energy to be released seismically. Consequently, secular crustal
strain rates estimated from GNSS data have been used to constrain
seismic hazard models such as UCERF3 \citep{Field2014}. Transient
crustal strain, which is caused by geophysical phenomena such as slow
slip events (SSEs) or postseismic deformation, is also relevant for
assessing seismic hazard. While transient strain itself is not
damaging, there is a risk that it can trigger major earthquakes
\citep{Roeloffs2006,Freed2001}. Dense networks of continuous GNSS
stations, such as the Plate Boundary Observatory (PBO), make it
possible to identify transient strain with high reliability. Developing
and improving upon methods for deriving secular and transient strain
from GNSS data is an active area of research.

% Existing methods for determining secular strain from GPS

Most methods for estimating strain rates from GNSS data assume some
parametric form for the deformation signal. The simplest method for
estimating secular strain rates assumes that GNSS derived velocities
can be described with a first-order polynomial (i.e, the deformation
gradients are constant) over some subnetwork of the GNSS stations
\citep[e.g.,][]{Feigl1990,Murray2000}. The components of the strain
rate tensor for each subnetwork are then determined from the least
squares fit to the observations. The assumption that deformation
gradients are spatially uniform is not appropriate when subnetworks
span a very large area. To help overcome this deficiency,
\citet{Shen1996,Shen2015} used an inverse distance weighting scheme,
in which the estimated strain rate at a point is primarily controlled
by observations at nearby stations. The method of
\citet{Shen1996,Shen2015} can be viewed as a form of local least
squares regression with a first-order polynomial \citep[e.g.,][sec.
6]{Hastie2009}. Other methods for estimating secular strain rates have
parameterized GNSS derived velocities with bi-cubic splines
\citep{Beavan2001}, spherical wavelets \citep{Tape2009}, and
elastostatic Green's functions \citep{Sandwell2016}. The type of basis
functions and the number of degrees of freedom for a parameterization,
which are often chosen subjectively, have a strong influence on the
strain solution. If there are too few degrees of freedom in the
parameterization, then estimated strain rates will be biased and the
uncertainties will be underestimated. On the other hand, if there are
too many degrees of freedom, then there will not be any coherent
features in the estimated strain rates. The methods described by
\citet{Beavan2001} and \citet{Tape2009} also require the user to
specify penalty parameters that control a similar trade-off between
bias and variance in the solution. One could parameterize deformation
with a physically motivated model of interseismic deformation
\citep[e.g.,][]{Meade2005,McCaffrey2007}. In such models the
lithospheric rheology and fault geometries are assumed to be known.
Any errors in the assumed physical model could result in biased strain
estimates and underestimated uncertainties.

@@

The temporally correlated noise in GNSS data has been thoroughly
studied over the past two decades \citep[e.g.,][]{Zhang1997, Mao1999,
Williams2004, Langbein2008} (See \citet{Bock2016} or \citet{He2017}
for a review of GNSS noise modelling). In these studies, temporally
correlated noise tends to be described with some combination of
Brownian motion (also known as random walk noise or a Weiner process),
a first-order Gauss-Markov (FOGM) process, and/or flicker noise. There
is some physical justification for using Brownian motion as a noise
model because it accurately describes the power spectrum of motion
resulting from instability in a geodetic monument
\citep[e.g.,][]{Wyatt1982, Wyatt1989}. In particular, the power
spectrums for Brownian motion and monument motion both decay
proportionally to $f^{-2}$, where $f$ is frequency. However, Brownian
motion necessarily contains a reference time at which the process
begins. Since there is no notion of when noise ``begins" in GNSS data,
we do not find Brownian motion to be an appropriate noise model. On
the other hand, a FOGM process has no reference time (i.e., it is
stationary), and its power spectrum, \begin{equation}\label{eq:FOGMPS}
S(f) = \frac{\beta^2}{(2 \pi f)^2 + \alpha^2}, \end{equation} decays
proportionally to $f^{-2}$ above the cutoff frequency $\alpha / (2
\pi)$. We then choose $z_i$ to be a spatially uncorrelated FOGM
process, which has the covariance function
\begin{equation}\label{eq:FOGM}
C_{z_i}\left((\pos,t),(\pos',t')\right) = \frac{\beta^2}{2\alpha}
\exp\left(-\alpha|t - t'|\right)\delta_{\pos,\pos'}, \end{equation}
where $\delta_{\pos,\pos'}$ is 1 if $\pos = \pos'$ and 0 otherwise.
See \citet[sec. B2]{Rasmussen2006} for more information on the FOGM
process.

>>>>

% Why we care about determining secular and transient strain

Crustal strain rates are fundamentally important quantities for
assessing seismic hazard. Knowing where and how quickly strain is
accumulating gives insight into where we can expect stored elastic
energy to be released seismically. Estimates of secular, or long-term,
strain rates have been used to constrain seismic hazard models such as
UCERF3 \citep{Field2014}. Transient crustal strain, which is caused by
geophysical processes such as slow slip events (SSEs) or postseismic
deformation, is also relevant for assessing seismic hazard. While
transient strain itself is not damaging, there is a risk that it can
trigger major earthquakes \citep{Roeloffs2006,Freed2001}.

% How secular strain rates are estimated from GNSS

Displacement data from dense GNSS networks, such as the Plate Boundary
Observatory (PBO), can be used to infer secular and transient strain.
Numerous methods for converting GNSS displacement data to strain have
been proposed, and most involve assuming some parametric form for the
deformation signal. The simplest method for estimating secular strain
rates assumes that GNSS derived velocities can be described with a
first-order polynomial (i.e, the deformation gradients are constant)
over some subnetwork of the GNSS stations \citep[e.g.,][]{Feigl1990,
Murray2000}. The components of the strain rate tensor for each
subnetwork are then determined from the least squares fit to the
observations. The assumption that deformation gradients are spatially
uniform is not appropriate when subnetworks span a very large area. To
help overcome this deficiency, \citet{Shen1996,Shen2015} used an
inverse distance weighting scheme, in which the estimated strain rate
at a point is primarily controlled by observations at nearby stations.
The method of \citet{Shen1996,Shen2015} can be viewed as a form of
local least squares regression with a first-order polynomial. Other
methods for estimating secular strain rates have parameterized GNSS
derived velocities with bi-cubic splines \citep{Beavan2001}, spherical
wavelets \citep{Tape2009}, and elastostatic Green's functions
\citep{Sandwell2016}. The type of basis functions and the number of
degrees of freedom for a parameterization, which are often chosen
subjectively, have a strong influence on the strain solution. If there
are too few degrees of freedom in the parameterization, then estimated
strain rates will be biased and the uncertainties will be
underestimated. On the other hand, if there are too many degrees of
freedom, then there will not be any coherent features in the estimated
strain rates. The methods described by \citet{Beavan2001} and
\citet{Tape2009} also require the user to specify penalty parameters
that control a similar trade-off between bias and variance in the
solution. One could parameterize deformation with a physically
motivated model of interseismic deformation \citep[e.g.,][]{Meade2005,
McCaffrey2007}. In such models the lithospheric rheology and fault
geometries are assumed to be known. Any errors in the assumed physical
model could result in biased strain estimates and underestimated
uncertainties.

@@

The temporally correlated noise in GNSS data has been thoroughly
studied over the past two decades (See \citet{Bock2016} or
\citet{He2017} for a review). Most studies found that the temporally
correlated noise can be described with Brownian motion (also known as
random walk noise or a Weiner process) or flicker noise
\citep[e.g.,][]{Zhang1997, Mao1999, Williams2004, Langbein2008,
Murray2017}. Less frequently used noise models include a band-pass
process \citep[e.g.,][]{Langbein2004} and a first-order Gauss-Markov
(FOGM) process \citep[e.g.,][]{Langbein2004,Dmitrieva2015}. There is
some physical justification for using Brownian motion as a noise model
because it accurately describes the power spectrum of motion resulting
from instability in a geodetic monument \citep[e.g.,][]{Wyatt1982,
Wyatt1989}. In particular, the power spectrums for Brownian motion and
monument motion both decay proportionally to $f^{-2}$, where $f$ is
frequency. However, Brownian motion necessarily contains a reference
time at which the process begins. Since there is no notion of when
noise ``begins" in GNSS data, we do not find Brownian motion to be an
appropriate noise model. On the other hand, a FOGM process has no
reference time (i.e., it is stationary), and its power spectrum,
\begin{equation}\label{eq:FOGMPS} 
S(f) = \frac{\beta^2}{(2 \pi f)^2 + \alpha^2}, 
\end{equation} 
decays proportionally to $f^{-2}$ above the cutoff frequency $\alpha /
(2 \pi)$. We choose $z_i$ to be a spatially uncorrelated FOGM process.
The covariance function for $z_i$ is then
\begin{equation}\label{eq:FOGM}
C_{z_i}\left((\pos,t),(\pos',t')\right) = \frac{\beta^2}{2\alpha} 
                                          \exp\left(-\alpha|t - t'|\right)\delta_{\pos,\pos'}, 
\end{equation}
where $\delta_{\pos,\pos'}$ is 1 if $\pos = \pos'$ and 0 otherwise
\citep[sec. B2]{Rasmussen2006}.
